Two Phase Debiased Calibration Simulation

### 2024-04-16 src/explore/20240411-tpdcsim.qmd
Updated prelim_paper.tex. Mostly finished with section 1.

$$
\begin{aligned}
X_{1i} &\stackrel{ind}{\sim} N(2, 1) \\
X_{2i} &\stackrel{ind}{\sim} Unif(0, 4) \\
X_{3i} &\stackrel{ind}{\sim} N(-1, 1) \\
\varepsilon_i &\stackrel{ind}{\sim} N(0, 1) \\
Y_{i} &= 3 X_{1i} + 2 X_{2i} + \varepsilon_i.
\end{aligned}
$$

@@

gen_pop <- function(obs) {
  x1 <- rnorm(obs, 2, 1)
  x2 <- runif(obs, 0, 4)
  x3 <- rnorm(obs, -1, 1)
  eps <- rnorm(obs)

@@

  return(tibble(X1 = x1, X2 = x2, X3 = x3, Y = y))
}

@@

  # Phase 1 
  p1_ind <- sample(1:nrow(pop_df), size = p1_obs)
  p1_df <- filter(pop_df, row_number() %in% p1_ind)

  # Phase 2
  p2_ind <- sample(1:nrow(p1_df), size = p2_obs)
  p2_df <- filter(p1_df, row_number() %in% p2_ind)

@@

# Each estimator should return two outcomes:
# 1. Point estimate, and
# 2. Variance estimate.
pi_star <- function(p1_df, p2_df, pop_obs, ...) {
  extra <- list(...)
  n1 <- nrow(p1_df)
  n2 <- nrow(p2_df)

  theta <- sum(p2_df$Y / n2)
  vhat <- (1 / n2 - 1 / pop_obs) * var(p2_df$Y)

@@

tp_reg <- function(p1_df, p2_df, pop_obs, ...) {
  extra <- list(...)
  n1 <- nrow(p1_df)
  n2 <- nrow(p2_df)

@@

  # Step 2: Predict
  theta <- mean(predict(mod, p1_df))

  # Step 3: Get variance estimator
  vhat <-
    (1 / n1 - 1 / pop_obs) * (mod$coefficients[2]^2 * var(p1_df$X1) + 
    mod$coefficients[3]^2 * var(p1_df$X2)) +
    (1 / n2 - 1 / n1) * var(p2_df$Y - predict(mod, p2_df))

@@

dc_poptots <- function(p1_df, p2_df, pop_obs, ...) {
  extra <- list(...)
  n1 <- nrow(p1_df)
  n2 <- nrow(p2_df)

@@

  samp_X <- select(p2_df, X1, X2) %>% as.matrix()
  samp_X <- cbind(rep(1, nrow(p2_df)), samp_X)
  pop_tot <- extra$pop_tot
  # FIXME: This does not work
  # TODO: Change everything to Poisson sample.
  dc_w <-  
    GEcalib(samp_X,
            d = rep(1, nrow(p2_df)),
            total = pop_tot,
            entropy = entropy,
            DS = TRUE)

@@

  Pi_dc <- diag(1 - n2 / pop_obs) * (dc_w / N_obs)^2
  z_dc <- model.matrix(~X1 + X2, data = p2_df)
  gam_dc <- solve(t(z_dc) %*% z_dc, t(z_dc) %*% samp_df$Y)
  vhat <- 
  (p2_df$Y - t(gam_dc) %*% t(z_dc)) %*% Pi_dc %*% t(p2_df$Y - t(gam_dc) %*% t(z_dc))

dc_esttots <- function(p1_df, p2_df, pop_obs, ...) {
  extra <- list(...)
  n1 <- nrow(p1_df)
  n2 <- nrow(p2_df)

@@

  pistar_est <- pi_star(p1_df, p2_df, nrow(pop_df))
  tpreg_est <- tp_reg(p1_df, p2_df, nrow(pop_df))
  dcpop_est <- 
    dc_poptots(p1_df, p2_df, nrow(pop_df),
               pop_tot = c(nrow(pop_df), sum(pop_df$X1), sum(pop_df$X2)))

  return(
    tibble(Est = c("pistar", "tpreg"),
           Theta = c(pistar_est[[1]], tpreg_est[[1]]),
           Var = c(pistar_est[[2]], tpreg_est[[2]]))
  )

@@

# Analyze Mean
mc_res %>%
  mutate(err = Theta - true_theta) %>%
  mutate(CI = abs(err) < qnorm(0.975) * sqrt(Var)) %>%
  group_by(Est) %>%
  summarize(bias = mean(err),
            rmse = sqrt(mean(err^2)),
            EmpCI = mean(CI),
            sdest = sd(Theta)) %>%
  ungroup() %>%
  mutate(Ttest = abs(bias) / sqrt(sdest^2 / B_sims)) %>%
  select(-sdest)

# Analyze Variance
mc_res %>%
  group_by(Est) %>%
  summarize(mcvar = var(Theta),
            estvar = mean(Var),
            varvar = var(Var)) %>%
  ungroup() %>%
  mutate(Ttest = abs(mcvar - estvar) / sqrt(varvar / B_sims))

>>>>

$$
\begin{aligned}
X_{1i} &\stackrel{ind}{\sim} N(2, 1) \\
X_{2i} &\stackrel{ind}{\sim} Unif(0, 4) \\
X_{3i} &\stackrel{ind}{\sim} N(0, 1) \\
X_{4i} &\stackrel{ind}{\sim} Unif(0.1, 0.9) \\
\varepsilon_i &\stackrel{ind}{\sim} N(0, 1) \\
Y_{i} &= 3 X_{1i} + 2 X_{2i} + \varepsilon_i \\
\pi_{1i} &= \Phi(-x_{3i} - 2)\\
\pi_{2i|1} &= x_{4i}.
\end{aligned}
$$

@@

gen_pop <- function(obs) {
  x1 <- rnorm(obs, 2, 1)
  x2 <- runif(obs, 0, 4)
  x3 <- rnorm(obs, 0, 1)
  x4 <- runif(obs, 0.1, 0.9)
  eps <- rnorm(obs)

@@

  return(tibble(X1 = x1, X2 = x2, X3 = x3, X4 = x4, Y = y))
}

@@

  # Phase 1 
  # Poisson sampling
  # mean(pi_i) = 0.1
  # ptemp <- pt(-pop_df$X3 - 2, 3)
  # pi_i <- ifelse(ptemp > 0.7, 0.7, ptemp)
  ptemp <- pt(-pop_df$X3, 3)
  pi_i <- ptemp
  del1 <- rbinom(nrow(pop_df), 1, pi_i)
  p1_df <- pop_df %>%
    mutate(pi1 = pi_i) %>%
    rename(pi2 = X4) %>%
    mutate(del1 = del1) %>%
    filter(del1 == 1)

  # SRS
  # p1_ind <- sample(1:nrow(pop_df), size = p1_obs)
  # p1_df <- filter(pop_df, row_number() %in% p1_ind)

  # Phase 2
  # del2 <- rbinom(nrow(p1_df), 1, p1_df$pi2)
  del2 <- rbinom(nrow(p1_df), 1, p1_df$pi2 / 2)

  p2_df <- p1_df %>%
    mutate(del2 = del2) %>%
    filter(del2 == 1)

  # SRS
  # p2_ind <- sample(1:nrow(p1_df), size = p2_obs)
  # p2_df <- filter(p1_df, row_number() %in% p2_ind)

@@

# Each estimator should return two outcomes:
# 1. Point estimate, and
# 2. Variance estimate.
pi_star <- function(p1_df, p2_df, pop_obs, ...) {
  extra <- list(...)

  theta <- sum(p2_df$Y / (p2_df$pi1 * p2_df$pi2)) / pop_obs

  # SRS Variance
  # vhat <- (1 / n2 - 1 / pop_obs) * var(p2_df$Y)

  # Poisson Variance
  pis <- p2_df$pi1 * p2_df$pi2
  vhat <- sum((1 / pis^2 - 1 / pis) * p2_df$Y^2) / pop_obs^2

@@

tp_reg <- function(p1_df, p2_df, pop_obs, ...) {
  extra <- list(...)

@@

  # Step 2: Predict
  # SRS
  # theta <- mean(predict(mod, p1_df))

  # Poisson
  theta <- (sum(predict(mod, p1_df) / p1_df$pi1) + 
  sum(1 / (p2_df$pi1 * p2_df$pi2) * (p2_df$Y - predict(mod, p2_df)))) / pop_obs

  # Step 3: Get variance estimator
  # SRS
  # vhat <-
  #   (1 / n1 - 1 / pop_obs) * (mod$coefficients[2]^2 * var(p1_df$X1) + 
  #   mod$coefficients[3]^2 * var(p1_df$X2)) +
  #   (1 / n2 - 1 / n1) * var(p2_df$Y - predict(mod, p2_df))

  # Poisson
  p12_df <- 
    left_join(p1_df, p2_df, by = join_by(X1, X2, X3, Y, pi1, pi2, del1)) %>%
    mutate(del2 = ifelse(is.na(del2), 0, 1))
  pred_12 <- predict(mod, p12_df)
  eta <- pred_12 + p12_df$del2 / p12_df$pi2 * (p12_df$Y - pred_12)

  v1 <- sum((1 - p12_df$pi1) / p12_df$pi1^2 * eta^2)
  v2 <- sum(1 / (p2_df$pi1 * p2_df$pi2) * 
           (1 / p2_df$pi2 - 1) * (p2_df$Y - predict(mod, p2_df))^2)

  vhat <- (v1 + v2) / pop_obs^2

@@

dc_poptots <- function(p1_df, p2_df, pop_obs, ...) {
  extra <- list(...)

@@

  d_vec <- 1 / (p2_df$pi2 * p2_df$pi1)
  if (entropy == "EL") {
    gdi <- -1 / d_vec
    gpinv <- d_vec^2

  } else if (entropy == "ET") {
    gdi <- log(d_vec)
    gpinv <- d_vec

  } else {
    stop("This type of entropy has not yet been implemented.")
  }

  samp_X <- cbind(rep(1, nrow(p2_df)), as.matrix(select(p2_df, X1, X2)))
  dc_w <-  
    GEcalib(cbind(samp_X, gdi),
            d = rep(1, nrow(p2_df)),
            extra$pop_tot,
            entropy = entropy,
            DS = FALSE)

@@

  Pi_dc <- diag(1 - 1 / d_vec) * (dc_w / pop_obs)^2
  z_dc <- model.matrix(~1 + p2_df$X1 + p2_df$X2 + gdi)
  gam_dc <- 
    solve(t(z_dc) %*% diag(gpinv) %*% z_dc, t(z_dc) %*% diag(gpinv) %*% p2_df$Y)
  vhat <- 
    (p2_df$Y - t(gam_dc) %*% t(z_dc)) %*% Pi_dc %*% t(p2_df$Y - t(gam_dc) %*% t(z_dc))

  return(list(theta = theta, Var = as.numeric(vhat)))
}

dc_esttots <- function(p1_df, p2_df, pop_obs, ...) {
  extra <- list(...)

  # Get estimates of population totals
  pop_est <- c(pop_obs,
               sum(p1_df$X1 / p1_df$pi1),
               sum(p1_df$X2 / p1_df$pi1),
               extra$pop_gdi)

  d_vec <- 1 / (p2_df$pi2 * p2_df$pi1)
  if (entropy == "EL") {
    gdi <- -1 / d_vec
    gpinv <- d_vec^2

  } else if (entropy == "ET") {
    gdi <- log(d_vec)
    gpinv <- d_vec

  } else {
    stop("This type of entropy has not yet been implemented.")
  }

  samp_X <- cbind(rep(1, nrow(p2_df)), as.matrix(select(p2_df, X1, X2)))
  dc_w <-  
    GEcalib(cbind(samp_X, gdi),
            d = rep(1, nrow(p2_df)),
            pop_est,
            entropy = entropy,
            DS = FALSE)

  # Variance Estimation
  ex_var_mat <- matrix(nrow = 2, ncol = 2)
  ex_var_mat[1, 1] <- sum((1 / p1_df$pi1^2 - 1 / p1_df$pi1) * p1_df$X1^2)
  ex_var_mat[1, 2] <- sum((1 / p1_df$pi1^2 - 1 / p1_df$pi1) * p1_df$X1 * p1_df$X2)
  ex_var_mat[2, 1] <- sum((1 / p1_df$pi1^2 - 1 / p1_df$pi1) * p1_df$X1 * p1_df$X2)
  ex_var_mat[2, 2] <- sum((1 / p1_df$pi1^2 - 1 / p1_df$pi1) * p1_df$X2^2)

  Pi_dc <- diag(1 - 1 / d_vec) * (dc_w / pop_obs)^2
  z_dc <- model.matrix(~1 + p2_df$X1 + p2_df$X2 + gdi)
  gam_dc <- 
    solve(t(z_dc) %*% diag(gpinv) %*% z_dc, t(z_dc) %*% diag(gpinv) %*% p2_df$Y)
  vhat <- 
  (p2_df$Y - t(gam_dc) %*% t(z_dc)) %*% Pi_dc %*% t(p2_df$Y - t(gam_dc) %*% t(z_dc)) +
  (gam_dc[2:3] %*% ex_var_mat %*% gam_dc[2:3] / pop_obs^2)

  return(list(theta = theta, Var = as.numeric(vhat)))
}

@@

  # ptemp <- pt(-pop_df$X3 - 2, 3)
  # pop_pi1 <- ifelse(ptemp > 0.7, 0.7, ptemp)
  ptemp <- pt(-pop_df$X3, 3)
  pop_pi1 <- ptemp

  pistar <- pi_star(p1_df, p2_df, nrow(pop_df), pop_df = pop_df)
  tpreg <- tp_reg(p1_df, p2_df, nrow(pop_df))
  dcpop <- 
    dc_poptots(
      p1_df,
      p2_df,
      nrow(pop_df),
      pop_tot = c(nrow(pop_df), sum(pop_df$X1), sum(pop_df$X2), sum(-pop_pi1))
    )
  dcest <-
    dc_esttots(
      p1_df,
      p2_df,
      nrow(pop_df),
      pop_gdi = sum(-pop_pi1)
    )

  return(
    tibble(Est = c("pistar", "tpreg", "dcpop", "dcest"),
           Theta = c(pistar[[1]], tpreg[[1]], dcpop[[1]], dcest[[1]]),
           Var = c(pistar[[2]], tpreg[[2]], dcpop[[2]], dcest[[2]]))
  )

@@

# Find any failed estimates
which(is.na(mc_res$Theta))
which(is.na(mc_res$Var))

# Analyze Mean
mc_res %>%
  mutate(err = Theta - true_theta) %>%
  mutate(CI = abs(err) < qnorm(0.975) * sqrt(Var)) %>%
  group_by(Est) %>%
  summarize(Bias = mean(err, na.rm = TRUE),
            RMSE = sqrt(mean(err^2, na.rm = TRUE)),
            EmpCI = mean(CI, na.rm = TRUE),
            sdest = sd(Theta, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Ttest = abs(Bias) / sqrt(sdest^2 / B_sims)) %>%
  select(-sdest) %>%
  mutate(Ind = case_when(
    Est == "dcest" ~ 4,
    Est == "dcpop" ~ 3,
    Est == "tpreg" ~ 2,
    Est == "pistar" ~ 1)
  ) %>%
  mutate(Est = case_when(
    Est == "dcest" ~ "DC-Est",
    Est == "dcpop" ~ "DC-Pop",
    Est == "tpreg" ~ "TP-Reg",
    Est == "pistar" ~ "$\\pi^*$")
  ) %>%
  arrange(Ind) %>%
  select(Est, Bias, RMSE, EmpCI, Ttest) %>%
  knitr::kable("latex", booktabs = TRUE, escape = FALSE, digits = 3) %>%
  cat(file = "tables/tpdcsim_mean.tex")

# Analyze Variance
mc_res %>%
  group_by(Est) %>%
  summarize(mcvar = var(Theta, na.rm = TRUE),
            estvar = mean(Var, na.rm = TRUE),
            varvar = var(Var, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Ttest = abs(mcvar - estvar) / sqrt(varvar / B_sims))
